### 2013-08-16 thesis-proposal/problem.tex
Updated thesis statement

\section{Introduction}

Jonathan Aldrich (ISR), chair
Brad A. Myers (HCII)
Robert Harper (CSD)
Premkumar Devanbu (UCSD)

Designing and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that simultaneously supports a natural programming style, helps users find and fix errors early, and consistently performs well across diverse application domains remains a grand challenge in computing. Indeed, these concerns are often at odds with one another, so system designers must make trade-offs. No small set of general-purpose constructs can completely satisfy every need. Evidence supporting this view comes from the large community of researchers and domain experts designing new domain-specific languages and tools, each tailored to the needs of a particular developer community or application domain.

%In implementing a new feature or abstraction, these experts can take one of several approaches. 
To realize a new construct, these experts often take a \emph{language-oriented approach}, developing and distributing it together with a new programming language. The flexibility of this approach, coupled with support from tools like compiler generators and language workbenches, has led to a proliferation of specialized languages, each built around a particular set of privileged constructs. Unfortunately, a key issue limiting the practicality of this approach occurs at the boundaries between languages -- specialized constructs particular to one language cannot always be safely and naturally expressed in another, so building a program out of components written in different specialized languages can often prove difficult.

@@

Stepping back, it is clear that a \emph{library-oriented approach}, shown in Figure \ref{approaches}b, where new constructs are implemented entirely within a single host language as libraries, should be preferred when possible.
% -- library-based constructs cannot interfere with one another (assuming a suitable namespacing standard), libraries benefit from common distribution infrastructure and tool support, and externally-accessible library interfaces need not be restricted to a fixed common subset of all available constructs (e.g. the object system of the JVM or CLI).  We diagram this key latter distinction in Figure \ref{approaches}. 
Of course, this is very often \emph{not} possible today because from the perspective of a library developer, the programming system is a largely inflexible, \emph{monolithic} entity. That is, the language's syntax, type checking rules and semantics are fixed in advance, the compiler is a black box implementation of these rules, and the tools operate according to domain-agnostic protocols that use only the declared structure of the program being developed as input. To enhance any of these aspects of the programming system, the expert must step out of the language, and often out of the programming system entirely.

@@

\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) With the language-oriented approach, novel constructs are packaged into separate languages. Users can only safely and naturally call into languages consisting of common constructs. (b) With the library-oriented approach, there is one language and novel constructs are packaged as normal libraries. Thus, interoperability is not a problem but in monolithic systems, this approach is less expressive.\label{approaches}}
\end{figure}

>>>>

\section{Introduction}\begin{figure}
\begin{center}
\includegraphics[scale=0.5]{approaches.pdf}
\end{center}
\vspace{-20px}
\caption{\small (a) With the language-oriented approach, novel constructs are packaged into separate languages. Users can only safely and naturally call into languages consisting of common constructs. (b) With the library-oriented approach, there is one language and novel constructs are packaged as normal libraries. Thus, interoperability is not a problem but in monolithic systems, this approach is less expressive.\label{approaches}}
\end{figure}
Designing and implementing a programming language together with its supporting tools (collectively, a \emph{programming system}) that has a sound theoretical foundation, supports a natural programming style, helps users identify and fix errors as early as possible, and performs well across diverse problem domains remains a grand challenge in computing. No small, fixed collection of abstractions and associated tools that can adequately address these criteria in all situations. Instead,  researchers and domain experts regularly design and implement novel domain-specific abstractions, provide alternative implementations of existing abstractions, and supplement these implementations with a variety of edit-time, compile-time and run-time behaviors in order to better satisfy these criteria within the constraints of particular application domains.

To realize a new abstraction or behavior, researchers and domain experts must take either a \emph{language-internal approach}, where they work entirely within an existing language and distribute their solutions as libraries, or a \emph{language-external approach}, where they either construct a new, distinct system or extend an existing system by some mechanism that is not part of the language itself (such as an extension mechanism supported by a particular compiler or development environment).

When it is possible, taking a \emph{language-internal approach} (aka a \emph{library-oriented approach}), where new constructs are implemented entirely within a single host language as libraries, is a far simpler and more practical approach. % -- library-based constructs cannot interfere with one another (assuming a suitable namespacing standard), libraries benefit from common distribution infrastructure and tool support, and externally-accessible library interfaces need not be restricted to a fixed common subset of all available constructs (e.g. the object system of the JVM or CLI).  We diagram this key latter distinction in Figure \ref{approaches}. 
Of course, this is very often \emph{not} possible today because from the perspective of a library developer, the programming system is a largely inflexible, \emph{monolithic} entity. That is, the language syntax and its static and dynamic semantics are fixed in advance, the compiler is a ``black box'' implementation of these fixed semantics, and other tools operate according to domain-agnostic protocols that use only the basic structure of the program being developed as input. To enhance any of these components of the programming system, the expert must step out of the language, and often out of the programming system entirely.

Thus, for the most sophisticated new constructs and behaviors, experts often prefer the  {language-oriented approach}, developing and distributing it together with a new programming language. The flexibility of this approach, coupled with support from tools like compiler generators and language workbenches, has led to a proliferation of specialized languages, each built around a particular set of privileged constructs. Unfortunately, a key issue limiting the practicality of this approach occurs at the boundaries between languages -- specialized constructs particular to one language cannot always be safely and naturally expressed in another, so building a program out of components written in different specialized languages can often prove difficult.

{\color{red} (Need to reorganize this bit a little bit)}

Although it is widely recognized that typed programming languages are well-positioned to achieve these goals, there is no small collection of types that can adequately satisfy these criteria in all cases. Instead, a large community of researchers and domain experts continue to design new domain-specific type systems and implement them in various programming systems, making trade-offs between these concerns as appropriate for the needs of the research or developer community they are targeting.

@@

@@

### 2013-08-16 thesis-proposal/timeline.tex
Updated thesis statement

Our plan for carrying out the work described above involves building significantly on existing work on Ace, as well as iterating on existing highly-simplified prototypes we have developed of all of the components described above.

In the first year, we will fully develop the theory of active types and completing development of the Ace language. We plan on releasing Ace to the broader community, as described in Section 5.1, within the coming year. In addition, we will implement initial versions of a core active code prediction system for Java.

In the second year, we will begin conducting capstone evaluations of Ace, further applying the theory to functional languages, and begin working on applications of active code prediction to code completion and rehabilitative programming systems.

In the third year, we will complete capstone evaluations of Ace, conduct capstone evaluations of active code prediction and the applications that we have developed, and begin prototyping an actively-typed structured editor, laying foundations for the long-term evolution of this research program beyond the scope of this grant.

Evaluations will consist of user studies, case studies, classroom studies and field studies based on instrumented versions of the tools that we develop. We plan on justifying not just the theoretical utility of active types, but their usefulness and usability in practice as well.

>>>>

\subsection{Spring 2013}
I will complete the work on Ace and active type theory and submit it for publication.

\subsection{Summer 2013}
I will complete the work on active code prediction and submit it for publication.

\subsection{Fall 2013}
I will finalize all work and write the final dissertation.

### 2013-08-16 thesis-proposal/work.tex
Updated thesis statement

\subsection{Syntax and Visual Representation}
Thus far, we have been discussing active types within the context of languages with a fixed syntax. However, in many specialized domains, a specialized syntax or visual representation for novel constructs may be appropriate. There are several approaches that we can take to include extensible syntax support in an actively-typed programming system. We have not settled on any one of these, so our proposed research will include a rigorous examination of the pros and cons of each approach.

\subsubsection{Extensible Parsing} 
Extensible parsers have been well-studied in the literature (e.g. \cite{journals/entcs/BrabrandSV03}), and such systems could be integrated with an actively-typed language without conflict with one caveat: each new syntactic form needs to have a dispatch protocol associated with it, to assign control over its semantics to one of its subexpressions. Such a mechanism will be one avenue we will evaluate for suitability.

\subsubsection{Flexible Reinterpretation}
An approach that we have prototyped, but not extensively utilized, in Ace is to allow active types to directly observe the syntax trees of subexpressions, rather than simply their types as we have described above. This allows a highly contextualized reinterpretation of a syntax tree, similar to capabilities found in macro systems (although without a direct rewriting component.) We will further explore this approach, particularly as we begin implementing abstractions that may not naturally fit within Python's existing syntactic forms in an intuitive way.

\subsubsection{Delimited Parsing}
A related alternative approach would be to partially parse a file, leaving delimited regions either unparsed or only tokenized and dispatching to an active type to do the actual parsing. This would give more flexibility to the active type, but also a greater responsibility to fully parse and give semantics to the provided source code. Nevertheless, this may be appropriate in certain cases where a highly-specialized local syntax is desirable for convenience (and flexible reinterpretation or active code completion at edit-time does not suffice.)

\subsubsection{Actively-Typed Structured Editors}
Projects such as Citrus \cite{Ko05} and Barista \cite{Ko06} are analogs of language frameworks for structured editors -- editors where source code elements are represented directly as a tree in a visual or semi-visual form, rather than written as a string as in conventional languages. These tools demonstrate that modular construction of such editors may be possible. An actively-typed structured editor would instead allow an active type to directly control the visual representation and interaction behaviors associated with expressions of that type. This is the most direct analog of active types in this domain, but requires a somewhat radical shift in program representation. It is unlikely that this can be integrated easily into existing languages and systems, as we have largely done in our other work. Thus, we will develop a separate prototype of this approach, and believe that the long-term evolution of actively-typed programming systems will be toward such a representation.

>>>>

%\subsection{Syntax and Visual Representation}
%Thus far, we have been discussing active types within the context of languages with a fixed syntax. However, in many specialized domains, a specialized syntax or visual representation for novel constructs may be appropriate. There are several approaches that we can take to include extensible syntax support in an actively-typed programming system. We have not settled on any one of these, so our proposed research will include a rigorous examination of the pros and cons of each approach.
%
%\subsubsection{Extensible Parsing} 
%Extensible parsers have been well-studied in the literature (e.g. \cite{journals/entcs/BrabrandSV03}), and such systems could be integrated with an actively-typed language without conflict with one caveat: each new syntactic form needs to have a dispatch protocol associated with it, to assign control over its semantics to one of its subexpressions. Such a mechanism will be one avenue we will evaluate for suitability.
%
%\subsubsection{Flexible Reinterpretation}
%An approach that we have prototyped, but not extensively utilized, in Ace is to allow active types to directly observe the syntax trees of subexpressions, rather than simply their types as we have described above. This allows a highly contextualized reinterpretation of a syntax tree, similar to capabilities found in macro systems (although without a direct rewriting component.) We will further explore this approach, particularly as we begin implementing abstractions that may not naturally fit within Python's existing syntactic forms in an intuitive way.
%
%\subsubsection{Delimited Parsing}
%A related alternative approach would be to partially parse a file, leaving delimited regions either unparsed or only tokenized and dispatching to an active type to do the actual parsing. This would give more flexibility to the active type, but also a greater responsibility to fully parse and give semantics to the provided source code. Nevertheless, this may be appropriate in certain cases where a highly-specialized local syntax is desirable for convenience (and flexible reinterpretation or active code completion at edit-time does not suffice.)
%
%\subsubsection{Actively-Typed Structured Editors}
%Projects such as Citrus \cite{Ko05} and Barista \cite{Ko06} are analogs of language frameworks for structured editors -- editors where source code elements are represented directly as a tree in a visual or semi-visual form, rather than written as a string as in conventional languages. These tools demonstrate that modular construction of such editors may be possible. An actively-typed structured editor would instead allow an active type to directly control the visual representation and interaction behaviors associated with expressions of that type. This is the most direct analog of active types in this domain, but requires a somewhat radical shift in program representation. It is unlikely that this can be integrated easily into existing languages and systems, as we have largely done in our other work. Thus, we will develop a separate prototype of this approach, and believe that the long-term evolution of actively-typed programming systems will be toward such a representation.
